ODGI: understanding pangenome graphs

### 2022-03-02 manuscript/main.tex
squeeze and rework the layout

The key data in a pangenome graph is a representation of the alignment or the homology relationships between the sequences represented.
Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to the \REVIEWED{[1,2] genomic sequences [1,2]} in the pangenome.
ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes. Such a a universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
%Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.

\REVIEWED{As a demonstration of this universality, we took the C4 graph and added gene annotation from GRCH38 to its nodes from a GFF file using \textit{odgi position} ($\S$\ref{sec:supp_gff}). The resulting TSV contains pairs of node and color. Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 gene region is highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}). Zooming to the nucleotide level, the annotation shows the single nucleotide differences of the \textit{C4A} and the \textit{C4B} genes (Fig.~\ref{fig:viz_extract_untangle-C4-bandage-zoom}).}

\textit{odgi position} can also translate graph and path positions between or within graphs, emitting the liftovers in BED format.
For a precise translation process when conversing a query position to a reference position in a repeat region, we apply the \textit{path jaccard} context mapping concept.
It could be that the found reference node is visited several times by the reference.
To ensure a precise translation, we select the reference position whose context (the multiset of $Node.id$s reached within a distance of e.g. 10kbp) has the best jaccard metric when compared to the query context. \REVIEWED{For a more detailed explanation of the \textit{path jaccard} concept please see $\S$\ref{sec:supp_path_jaccard}.}
%the best jaccard for the query step and each reference candidate step is calculated by looking at a context range (e.g. 10kbp) around the locus.

>>>>

The key data in a pangenome graph is a representation of the alignment (i.e., the homology relationships) between \REVIEWED{[1,2] genomic sequences [1,2]}.
Navigating and understanding the graph requires coordinate systems to link other data to the sequences represented in the graph model.
ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same sequences.
Such a a universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different sequences in the same or different graphs.
%Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
\REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCH38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}). The resulting TSV contains pairs of node and color. Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 genes are highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}). Zooming to the nucleotide level, the annotation shows the single nucleotide differences of the \textit{C4A} and the \textit{C4B} genes (Fig.~\ref{fig:viz_extract_untangle-C4-bandage-zoom}).}

\textit{odgi position} can also translate graph and path positions between or within graphs, emitting the liftovers in BED format.
For a precise translation process when conversing a query position to a reference position in a repeat region, we apply the \textit{path jaccard} context mapping concept.
It could be that the found reference node is visited several times by the reference.
To ensure a precise translation, we select the reference position whose context (the multiset of $Node.id$s reached within a distance of e.g. 10kbp) has the best jaccard metric when compared to the query context. \REVIEWED{For a more detailed explanation of the \textit{path jaccard} concept see $\S$\ref{sec:supp_path_jaccard}.}
%the best jaccard for the query step and each reference candidate step is calculated by looking at a context range (e.g. 10kbp) around the locus.
